### 2016-07-29 text/dsltrans.tex
Replace DSLTrans section with Frankenstein from different sources.

\section{The DSLTrans Transformation Language}
\label{sec:dsltrans}

@@

@@

\begin{figure*}[t]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/organization.pdf}
                \caption{Organization language}
                \label{fig:OrganizationLanguage}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/gender.pdf}
                \caption{Gender language}
                \label{fig:GenderLanguage}
        \end{subfigure}%
        \caption{Metamodels for the Police Station transformation}
        \label{fig:squadmetamodel}
\end{figure*}

@@

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.30\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_hierarchy.pdf}
                \caption{Original model}
                \label{fig:police_hierarchy}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_gender.pdf}
                \caption{Transformed model}
                \label{fig:police_gender}
        \end{subfigure}%
        \caption{Model before and after transformation}
        \label{fig:transformationexample}
\end{figure*}
\reviewer{Figure 3: I would suggest adding another node f5 superviesd by f4, to illustrate the
notion of transitive closure. Another remark: I think the black containment dia-
monds do not belong in an instance model, as they are a metamodel notion.}

@@

\subsection{Properties to Prove}

The properties we aim to prove on the Police Station transformation are structural properties. These properties are composed of a pre-condition and a post-condition component, as seen in \cref{fig:properties}.

Informally, a property can be read as `if the pre-condition graph matches on the input model to the transformation, then the post-condition graph will match any output model produced'. Further details as well as formal validity and completeness of the property proving process are discussed in \cref{sec:verif_dsltrans_props}.

As a brief example of property syntax and semantics, consider the property in \cref{fig:dsltrans_prop1}. The pre-condition graph is composed of a Station element connected to a Female element and a Male element, where all elements are from the \emph{Organization language} metamodel. This structure is repeated in the post-condition graph, with the difference that the metamodel for these elements is the \emph{Gender language}. Thus, this property represents the statement ``\emph{a model which includes a
police station that has both male and female officers will be
transformed into a model where the male officer will exist in the male set
and the female officer will exist in the female set}''. We expect this property to always hold in our transformation.

\begin{figure*}[thb]
        \centering
        \begin{subfigure}[b]{0.45\textwidth}
                \centering
                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property1.pdf}
                \caption{Property 1 -- Expected to hold}
                \label{fig:dsltrans_prop1}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.45\textwidth}
                \centering
                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property2.pdf}
                \caption{Property 2 -- Not expected to hold}
                \label{fig:dsltrans_prop2}
        \end{subfigure}%
        \caption{ Properties to be proved on the Police Station transformation}
        \label{fig:properties}
\end{figure*}

\reviewer{Figure 4: The two “agent” edges in the postcondition should be marked “female”
resp. “male”} 
In contrast, we do not expect the property in \cref{fig:dsltrans_prop2} to always hold. This property represents the statement ``\emph{any model which includes a
female officer will be transformed into a model where that female officer will
always supervise another female officer}''. It is not difficult to construct an input model where the pre-condition holds, but the post-condition does not. For example, take an input model that contains only one female officer, as there will only be one female officer in the output model.

We shall discuss our property proving technique in \cref{sec:verif_dsltrans_props}. Following this, experiments in \cref{sec:experiments} will present experimental results from proving these two properties on the Police Station transformation.

@@

>>>>

\section{DSLTrans}
\label{sec:DSLTrans_formal}

@@

\subsection{DSLTrans Introduction}

DSLTrans is a visual graph-based and rule-based model
transformation engine that has two important properties enforced by construction: all its computations
are both \emph{terminating} and \emph{confluent}~\cite{Barroca2011}.
These properties stem from the fact that DSLTrans does not allow
unbounded loops during execution, making it a Turing-incomplete
computing language~\cite{Barroca2011}. Besides their obvious importance in
practice, \emph{termination} and \emph{confluence} were instrumental in the
implementation of our verification technique for pre-/post-condition contracts.

Model transformations are expressed in DSLTrans as sets of graph rewriting
rules, having an upper part (named \emph{MatchModel}), a lower part (\emph{ApplyModel}) and,
optionally, negative application conditions. The main construction
used in the scheduling of model transformation rules in DSLTrans is a
\emph{layer}. Each model transformation rule in a layer cannot match over the output of any other rule in the same layer. As well, rules cannot modify the input graph during the rewriting phase (termed \emph{out-place} execution). Layers are organized sequentially and
the output model that results from executing a given layer is passed as input to
the next layer in the sequence.

%Model transformations are expressed in DSLTrans as sets of graph rewriting
%rules, having the classical left-hand side (LHS), right-hand side (RHS) and,
%optionally, negative application conditions. The main construction
%used in the scheduling of model transformation rules in DSLTrans is a
%\emph{layer}: a layer contains a set of model transformation rules that execute
%without being able to see each other's output. This forces the rules in a layer
%to execute independently from each other. Layers are organized sequentially and
%the output model that results from executing a given layer is passed as input to
%the next layer in the sequence.

A DSLTrans rule can match over the elements of the input model of the
transformation and also over elements that have been generated so far in the
output model. Matching over elements of the output model of a
transformation is achieved using a DSLTrans construct called \emph{backward
links}. Backward links allow matching over traces between elements in the input and
the output models of the transformation. These traces are explicitly built by
the DSLTrans transformation engine during rule execution.

 \begin{figure}[t]
   \begin{center}
     \includegraphics[width=0.40\textwidth]{figures/DSLTrans/UWRule.pdf}
     \caption{An example of a DSLTrans rule}
     \label{fig:DSLTrans_rule}
   \end{center}
   \vspace{-0.25in}
 \end{figure}

For example, we depict in Figure~\ref{fig:DSLTrans_rule} a
rule in the DSLTrans language.
When a rule is executed, the graph in the \emph{MatchModel} of the rule is searched for in the transformation's input model, together with the classes in the \emph{ApplyModel} of the rule that are connected to \emph{backward
links}. An example of a \emph{backward link} can be observed in
Figure~\ref{fig:DSLTrans_rule} as a dotted line connecting the \emph{Country} and the
\emph{Community} match classes. During the rewrite part of rule application,
the instances of classes in the \emph{ApplyModel} of the rule that are not connected to
backward links, together with their adjacent relations, are created in the
output model.

For example, the \emph{UnionWomanRule} rule in Figure~\ref{fig:DSLTrans_rule} will match over a \emph{Country} element connected to a \emph{Family} element connected to a \emph{Parent} element. If these elements are found in the input model along with the corresponding \emph{Community} and \emph{Woman} elements in the output model, then a \emph{persons} relation will be created between those output elements.

Although not present in this rule,
copying object attribute values from the \emph{MatchModel} to the \emph{ApplyModel} of the rules is
also part of the DSLTrans language, as illustrated in Section~\ref{sec:DSLTransRepresentation}.

@@

%\begin{figure*}[t]
%        \centering
%        \begin{subfigure}[b]{0.40\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/organization.pdf}
%                \caption{Organization language}
%                \label{fig:OrganizationLanguage}
%        \end{subfigure}%
%        ~~
%        \begin{subfigure}[b]{0.40\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/gender.pdf}
%                \caption{Gender language}
%                \label{fig:GenderLanguage}
%        \end{subfigure}%
%        \caption{Metamodels for the Police Station transformation}
%        \label{fig:squadmetamodel}
%\end{figure*}

@@

%\begin{figure*}[htb]
%        \centering
%        \begin{subfigure}[b]{0.30\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_hierarchy.pdf}
%                \caption{Original model}
%                \label{fig:police_hierarchy}
%        \end{subfigure}%
%        ~~
%        \begin{subfigure}[b]{0.40\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_gender.pdf}
%                \caption{Transformed model}
%                \label{fig:police_gender}
%        \end{subfigure}%
%        \caption{Model before and after transformation}
%        \label{fig:transformationexample}
%\end{figure*}
\reviewer{Figure 3: I would suggest adding another node f5 superviesd by f4, to illustrate the
notion of transitive closure. Another remark: I think the black containment dia-
monds do not belong in an instance model, as they are a metamodel notion.}

@@

In addition to the constructs presented in the example in
Figure~\ref{fig:DSLTrans_rule}, DSLTrans has several others:
\emph{existential matching} which allows selecting only one result when a match class of a rule
matches an input model, \emph{indirect links} for transitive matching
over containment relations in the input model, and \emph{negative application
conditions} that allow the transformation designer to specify conditions under
which a rule should not match. These constructs are not currently used in
our verification approach, and the interested reader is referred to~\cite{Barroca2011} for further information.

@@

\subsection{Families To Persons}

As our running example we present an extended version of the \FTP
transformation described in \cite{Oakes2015}. The original \FTP transformation can be
found in the ATL zoo~\cite{atlZoo}, and has also been discussed in a number of related works
on verification and testing~\cite{Gogolla2011}.

We chose this \FTP transformation as our running example for two reasons. First, it transforms domains whose concepts are easily understandable by anyone (cf. Section~\ref{sec:trans_domain}).
Second, it has a certain degree of complexity since it uses many features available in the ATL language (cf. Section~\ref{sec:trans_explanation}).

\subsection{Transformation Domains}\label{sec:trans_domain}

\begin{figure}[bt]
\centering
\subfigure[\emph{Families\_Extended} metamodel]{
\includegraphics[width=0.97\columnwidth]{figures/DSLTrans/Families_Extended.pdf}
\label{fig:fams}
}
\subfigure[\emph{Persons\_Extended} metamodel]{
\includegraphics[width=0.97\columnwidth]{figures/DSLTrans/Persons_Extended.pdf}
\label{fig:pers}
}
\caption{Metamodels of the \emph{Families-to-Persons\_Extended} transformation}\label{fig:metamodels}
\end{figure}

The input and output metamodels of this transformation are shown in Figure~\ref{fig:metamodels}. Please note that abstract classes are depicted in grey color and with italic names, and inheritance relationships are depicted in grey.

The input metamodel, the \emph{Families\_Extended} metamodel, has the \emph{Country} class as a root element.
A \emph{Country} is made up of \emph{companies}, \emph{families} and \emph{cities}.
A \emph{Family} has a \emph{lastName}, is \emph{registeredIn} a \emph{Neighborhood} and can have any number of \emph{mothers} and \emph{fathers}, who are \emph{Parent}s and may, in turn, work in (\emph{worksIn}) a \emph{Company}.
It can also contain any number of \emph{sons} and \emph{daughters}, who are \emph{Child}ren, and every child \emph{goesTo} a \emph{School}.
Both parents and children are \emph{Member}s that have a \emph{firstName}, belong to a \emph{family} and each of them \emph{livesIn} a \emph{City}.

A \emph{City} may contain \emph{companies}, and a \emph{Company}, in turn, can be present in (relationship \emph{isIn}) several distinct cities.
A \emph{City} is composed of \emph{neighborhoods}, and these can have \emph{schools}, where several \emph{students} are registered.
Every \emph{School} has \emph{Service}s, and these may be \emph{special}, for students with special needs, or simply offer \emph{ordinary} services.
Finally, countries, cities, companies, neighborhoods and schools have a \emph{name} attribute, which is inherited from the abstract \emph{NamedElement} class.

The output metamodel, \emph{Persons\_Extended}, is shown in Figure~\ref{fig:pers}.
The root class is \emph{Community}, which is made up of \emph{persons}, \emph{townHalls} and \emph{associations}.
A \emph{Person} has a \emph{fullName} and can either be a \emph{Man} or a \emph{Woman}.
An \emph{Association} has a \emph{Committee} that makes decisions.
Every \emph{TownHall} has a roster of \emph{workers} (all the persons that are employed), hosts a \emph{Committee} to make decisions, and also governs several \emph{districts}.
A \emph{District} may contain several \emph{facilities}, either of type \emph{SpecialFacility} for those with special needs, or \emph{OrdinaryFacility}.
Each \emph{Facility} may have registered several persons as \emph{members}.
Finally, associations, town halls, committees, districts and facilities have a \emph{name} attribute.

\subsection{DSLTrans Representation}\label{sec:DSLTransRepresentation}

\begin{figure*}
  \begin{center}
  \includegraphics[width=0.94\textwidth]{figures/DSLTrans/DSLTrans_Rules_javi.pdf}
  \caption{DSLTrans version of the \emph{Families-to-Persons\_Extended} transformation}
  \label{fig:DSLTrans_rules}
  \end{center}
\end{figure*}

Figure~\ref{fig:DSLTrans_rules} displays the DSLTrans transformation which corresponds to the ATL \emph{Families-to-Persons\_Extended}\\
transformation shown in Listing~\ref{lst:F2P_Extended}. Let us mention here that we have removed five rules from the figure to improve visual clarity.
There is a vertical dotted blue line for each of these rules, located where the rules have been removed. The missed rules are similar to those that surround them, and can therefore be safely ignored in our explanation.

%As explained in Section~\ref{sec:dsltrans}, each DSLTrans rule has a \emph{MatchModel} and an \emph{ApplyModel}.
%As we can see, each of them has a left-hand side (LHS) and a right-hand side (RHS)\levi{not sure about the LHS/RHS reference here\ldots}.
%There are elements of a specific type and associations among them displayed in
% both the upper part (named \emph{MatchedModel} in DSLTrans) and lower part (called \emph{ApplyModel}) of the rules.
%In each side, elements of a specific type and relationships among them are displayed.
%Elements can also have attributes.
%Rule \emph{Households2Community} corresponds to rule R1 in
%Listing~\ref{lst:Families2Persons}, \emph{Father2Man} corresponds to rule R2 and \emph{UnionFather}
%to binding B11.

%They appear in the \emph{ApplyModel} when they have to be initialized, and
%in the \emph{MatchedModel} when they are used to initialize attributes in the \emph{ApplyModel}.
%For instance, attributes \emph{fullName} and \emph{lastName} of \emph{Member} and \emph{Family} in rule \emph{Father2Man} are used
%to initialize attribute \emph{fullName} of \emph{Man}.
%%
%%Other than these attributes, there are others in the RHS which appear in an element
%%whenever such element is going to be used in backward links, and they also appear when they are being used in such links. These attributes are always referred to with the name \emph{ApplyAttribute}.
%%For instance, we can see in rule \emph{Households2Community}, which is created for R1 in Listing~\ref{lst:Families2Persons}, that a \emph{Community} is created for a \emph{Households}.
%%Besides, since these two elements are going to be used later in a backward link, the \emph{ApplyAttribute} attribute is set in \emph{Community}. Note that the \emph{ApplyAttribute} attribute of an element must be created when the element is created, and the rule where it is created must be in the sequence of rules before the rule that uses the element in a backward link.
%
%Rule \emph{Households2Community} corresponds to R1 in
%Listing~\ref{lst:Families2Persons}, although the reference \emph{has} is not
%initialized in this rule.
%
%Rule \emph{Father2Man} corresponds to R2. We can see how the two elements in the
%\emph{from} part of the ATL rule appear in the \emph{MatchedModel} of the DSLTrans rule, as well
%as the filter used in ATL, which is represented as an association. The
%element \emph{Man} appears in the \emph{ApplyModel}, and its \emph{fullName} attribute is
%updated with the concatenation of the \emph{firstName} and \emph{lastName}
%attributes. A similar rule, not shown here, creates elements of type
%\emph{Woman}, which corresponds to R3 in Listing~\ref{lst:Families2Persons}.
%
%The last rule is used to set the \emph{has} relation between a \emph{Community}
%and a \emph{Man}. Here, backward links are used. The first one is to indicate
%that the \emph{Community} element has been created from a \emph{Households}
%element (which was done in rule \emph{Households2Community}). The other two
%backward links state that the \emph{Man} is created from a \emph{Family} and a
%\emph{Member} that are linked with the relation \emph{father} (such as in rule
%\emph{Father2Man}). This rule corresponds to binding B11 in
%Listing~\ref{lst:Families2Persons}. There is a similar rule, not shown in the
%figure, for initializing the \emph{has} association between \emph{Community} and
%\emph{Woman}, what corresponds to B12.

The process of constructing a DSLTrans transformation from an ATL one is described in the next section.
For now, note that DSLTrans transformation obtained from ATL through the higher-order transformation includes only one rule per layer, meaning all rules execute sequentially.
This is due to the sequential semantics of ATL that we replicate in DSLTrans.

Also, note that attribute copies are represented with arrows from the \emph{ApplyModel} of the rule to the \emph{MatchModel}, such as in rule \emph{Neighborhood2District}, where the created \emph{District} gets the same name as the matched \emph{Neighborhood}.
The string of an attribute of a created element can also be initialized with the concatenation of several strings.
For instance, in rule \emph{Father2Man}, the full name of the created \emph{Man} comes from the concatenation of the first name of the matched \emph{Parent} and the last name of his/her \emph{Family}.
Or it can be assigned the string of an attribute of an element in the \emph{MatchModel} concatenated with a given string, such as in rule \emph{City2TownHall}.

%\emph{Father2Man} in Figure~\ref{fig:DSLTrans_transformation} performs an attribute
%copy from the match to the apply part of the rule, graphically represented with arrows from the \emph{ApplyModel} to the \emph{MatchModel}. More precisely, it
%fills in the \emph{fullName} attribute of the generated \emph{Man} instance by
%concatenating the \emph{firstName} and \emph{lastName} attribute values found
%respectively in the \emph{member} and \emph{family} instances identified by the
%match part of the rule. This operation replicates the concatenation of the
%\emph{firstName} and the \emph{lastName} occurring in the rule component marked
%\emph{B2} in Listing~\ref{lst:Families2Persons}.

%The first three rules in the DSLTrans transformation replicate the R1, R2, and R3
% portions of the ATL transformation. Then, the last four rules in
% Figure~\ref{fig:DSLTrans_transformation} connect those elements and set the values, replicating B11, B12, B2, and B3 in Listing~\ref{lst:Families2Persons}.
%
% This transformation has been built automatically from its ATL
%counterpart using a higher-order transformation, which will be presented in Section~\ref{sec:mapping}. For now, we will briefly the differences and
%similarities between the two transformations:
%
%\begin{itemize}
%  \item The DSLTrans F2P transformation is completely graphical, while the ATL
%  F2P transformation is fully textual.
%  \item The DSLTrans F2P is composed of layers that are organized sequentially.
%  A layer in DSLTrans is a box (e.g. the \emph{Root} rule in
%  Figure~\ref{fig:DSLTrans_transformation}) that can contain one or more rules.
%  In the transformation in Figure~\ref{fig:DSLTrans_transformation}) each layer
%  contains only one rule, as that layout is automatically produced by the higher
%  order transformation.
%  \item The layers belonging to the transformation in
%  Figure~\ref{fig:DSLTrans_transformation} execute one after the other, passing
%  the result of one layer to the next one.
%\end{itemize}

\subsection{Formal Structures}

This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.

As discussed in Section~\ref{subsec:DSLTrans_syntax}, DSLTrans transformations are composed of rules arranged in layers.

\subsubsection{Similar Structures}

Note that in the definitions that follow, in this section and others, that we define structures which are very similar in composition, which each containing two or three typed graphs, as well as a link component. This similarity is essential to our technique, as we will define morphisms between the various components of these constructions, which vary depending on the particular structure under examination.

While this composition of typed graph approach is preferred by the authors due to its reflection in our prover implementation, some readers may disagree. In this case, note that it is equivalent to think of these structures as one large typed graph annotated by the component it originates from, with appropriate projections for selecting relevant portions.

\subsubsection*{DSLTrans Transformation Rule}

A transformation rule is the elemental block of a DSLTrans transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.

A transformation rule includes a non-empty match pattern and a non-empty apply pattern. This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}. As described in Section 3.4 or Section 2, which the match graph elements are found in the input model, then the apply graph elements are produced in the output model.

A match pattern can include indirect links that are used to transitively match containment relations in a model. The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match. An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.

A rule may also contain a negative application condition (NAC) which if matched over the target graph prevents application of the rule.

A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}. As described in that section, backward links define dependencies between rules. \bentley{Expand}

\begin{definition}{DSLTrans Transformation Rule\\}
\label{def:transformation_rule}

A DSLTrans transformation rule is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:

\begin{itemize}
\item $\mathit{NAC}$, $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
\item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $

\end{itemize}

Note that when we require an element of the $\mathit{Match}$, $\mathit{Apply}$, or $\mathit{NAC}$ graphs, such as the vertices, we will index the required element. For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.

\begin{itemize}
\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
\item $E_{back}$ contains the backward links
\begin{itemize}
\item $E_{back} $ is disjoint from $E_{Match}$,  $E_{Apply}$, and $E_{NAC}$
\end{itemize}
\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the pattern and target vertices for each backward link
\begin{itemize}
\item Note the source of backward links is a vertex in $V_{\textit{Apply}}$ while the target is a vertex in $V_{\textit{Match}}$
\end{itemize}
\end{itemize}

\textsc{Rules} is the set of all rules.

\end{definition}

We additionally impose that for the well-formedness of a DSLTrans rule, there always exists an element or edge to be created in the \textit{Apply} graph to be created. That is, either there are edges to be created in the \textit{Apply} graph, or there is a vertex in the \textit{Apply} graph that is not the source of a backward link.

\begin{itemize}
\item $E_{\textit{Apply}} \neq \emptyset \lor$
\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s_{back}(e) \neq v \big\}$
\end{itemize}

\subsubsection*{DSLTrans Layer and Transformation}

\cref{def:layer} and \cref{def:transformation} formalise the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}. An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation. As expected, a DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.

\begin{definition}{Layer\\}
\label{def:layer}
A layer is a finite set of transformation rules: $\{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$.

The set of all layers is denoted $\textsc{Layers}$. 
\end{definition}

Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution. As discussed in Section BLAH, these semantics ensure that rules are independent, and cannot act on the output of other rules within the same layer.

\begin{definition}{DSLTrans Transformation\\}
\label{def:transformation}

A DSLTrans transformation is a finite list of layers denoted $\{l_0, l_1, \dots, l_n | l \in \textsc{Layers}\}$. Note that the order of layers in a transformation is important.

The set of all transformations is denoted $\textsc{Transforms}$.

\end{definition}

\subsubsection*{Input-Output Model}

To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct. This input-output model allows the representation of the initial state as well as intermediate operational states during the execution of a model transformation. These input-output models are thus the `path conditions' which represent all transformation executions through an abstraction relation. This is further discussed in Section ABSTRACTION RELATION.

The structure of a input-output model is intentionally very similar to a DSLTrans rule. There is one typed graph representing the input model and another typed graph representing the output model. As well, the construct contains a set of edges, named \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.

\begin{definition}{Input-Output Model\\}
\label{def:input_output_model}

An input-output model rule is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:

\begin{itemize}
\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
\begin{itemize}
\item $E_{trace}$ contains the traceability links
\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
\end{itemize}

\end{itemize}

Let $\mathit{IOM}$ be the set of all input-output models.

We define a utility function getTransformation: IOM $\rightarrow$ Transforms. This function returns the Transform that the input-output model was built for. The purpose of this function is to restrict input-output models to only be applicable for the transformation they represent.

Note that the $\VT$ and $\ET$ for input and output will come from the input and output metamodels. \bentley{Expand}

\end{definition}

\subsection{Transformation Semantics}
\label{sec:DSLTrans_semantics}

This section will discuss the semantics of a DSLTrans transformation. Given an input model and a transformation, an output layer will be produced through the repeated execution of rules and layers.

\subsubsection*{Execution of a DSLTrans Rule}

We will now address the execution of a rule in the DSLTrans language, basing our explanation on double pushouts.

Note that we will apply rules differently, depending on whether they contain indirect links or not. We will begin with the simpler case where the rule does not contain indirect links, diagrammed in Figure~\ref{fig:dpo}.

We create first the matcher and the rewriter for a DSLTrans rule.

\subsubsection*{Matcher of a Transformation Rule}

To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.

Note that the rule's NAC will also be considered in the matching of a rule.

\begin{definition}{Matcher of a Transformation Rule\\}
\label{def:matcher_transformation_rule}

Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.

We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
\begin{itemize}
\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
\item $E = E_{Match} \cup E_{back}$
\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
\item $\tau = \tau_{Match} \cup \tau_{Apply}$
\item $\VT = \VT_{Match} \cup \VT_{Apply}$
\item $\ET = \ET_{Match} \cup \ET_{Apply}$
\end{itemize}

\end{definition}

\subsubsection*{Rewriter of a Transformation Rule}

To continue with the double pushout approach, we must construct the rewriting (or replacement) graph.

The construction of this rewriter is essentially the same as the underlying rule. However, to support traceability, we require that backward links in the rule be converted into traceability links, and that new traceability links be created between all match and apply vertices.

\begin{definition}{Rewriter of a Transformation Rule\\}
\label{def:rewriter_transformation_rule}

Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.

We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a structure $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:

\begin{itemize}
\item $ \mathit{Input} = \mathit{Match}$
\item $ \mathit{Output} = \mathit{Apply}$
\end{itemize}

We then modify $\mathit{trace}$ to contain the appropriate traceability links

\begin{itemize}
\item $E_{trace} = \exists e \in E_{trace} | $
\begin{itemize}
\item $ \forall m \in V_{Match}, a \in V_{Apply}:$$s_{trace}(e) = a$, $t_{trace}(e) = m$ 
\end{itemize}
\end{itemize}

\end{definition}

\subsubsection*{Double-Pushout Approach}

\begin{figure}[h!] \centering
\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC}
	\caption{Double pushout approach}
	\label{fig:dpo}
\end{figure}

If the reader is unfamiliar with the theory of double-pushouts, we refer to work on Fundamentals of Algebraic Graph Transformation.

The core idea of double-pushouts is to present graph rewriting in terms of morphisms, which compartmentalize the actions of finding a match and adding/deleting edges.

The application of a rule is called a \textit{production}. Below, we will first explain the production of a single DSLTrans rule, expand this production to handle indirect links in the rule, then discuss multiple productions.

Let this production be for a rule $r = \langle \mathit{NAC}, \mathit{[Match}, \mathit{Apply}, \mathit{back} \rangle$. To simplify the explanation, we constrain our rules to not include any indirect links. That is, $\nexists E \in E_{\textit{Match}} \cup E_{\mathit{NAC}} | \mathit{isIndirect}(E)$.

Figure~\ref{fig:dpo} presents a basic diagram for the approach.
The major components are:
\begin{itemize}
\item L: The matcher of the rule $\lceil r \rceil$
\item R: The rewriter of the rule $\lfloor r \rfloor$
\item G: The graph to be matched/rewritten
\item H: The rewritten graph
\end{itemize}

\paragraph{Rule Components}
K is the interface graph between L and R. That is, elements in K appear in both L and R. Elements which are to be deleted by the production appear in L but not K, and elements to be created are present in R but not K.

In the absence of indirect links, we note that $K = L$ as elements cannot be deleted in DSLTrans rules. Therefore the morphism l is isomorphism.

The morphism r from K to R is an injective typed graph homomorphism. Note that as described in Section~\ref{subsubsec:structure_homomorphism}, this homomorphism will be composed of sub-homomorphisms between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.

N represents the typed graph NAC of the rule. We note that this production cannot occur if there is a typed graph edge homomorphism c between the rule's NAC and G.

\paragraph{Graph Components}

G is the target graph which is to be matched on.

m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model.

D is the interface graph between the original graph G, and the rewritten graph H. Similar to K, D contains elements in both. Again, as DSLTrans rules do not delete elements, D will be isomorphic to G in this case, and f will be isomorphism. The morphism k will embed the K interface graph within the interface graph D.

The morphism g from the interface graph D to the rewritten graph H gives the graph elements, while the morphism $r$ gives the embedding of the rewriter R with the elements to be produced. Thus, it is the pushout of R and D through K that gives H. g and n will be typed graph morphisms.

\subsubsection{Double-Pushout Approach with Indirect Links}

If the DSLTrans rule contains indirect links, we must modify the double-pushout approach described above to handle the transitive closure that the construct implies. The main change is to build the transitive edges, match on them, and then not build them in the final graph.

Note that this second explanation is valid for rules containing indirect links, or not containing them.

\begin{figure}[h!] \centering
\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC_and_indirect}
	\caption{Double pushout approach with indirect links}
	\label{fig:dpo_indirect}
\end{figure}

We add the component G' to the diagram, which represents the transitive closure of the Input graph of G. The homomorphism i gives the embedding of G into G'.

The homomorphism m will try to map the indirect links onto the newly created edges in G'.

K will not contain the indirect links. Therefore l will not longer be surjective, as nothing will map to the indirect links. R and r remain the same.

The interface D does not contain the edges newly created in G'. Thus the morphism f does not map onto these edges in G'.

n, g, and H remain the same.

\paragraph{Example}

\begin{figure}[h!] \centering
\includegraphics[width=0.44\textwidth]{figures/formal/dpo_indirect_example}
	\caption{DPO example with indirect links}
	\label{fig:dpo_indirect_example}
\end{figure}

Figure~\ref{fig:dpo_indirect_example} shows an example of the DPO approach when the rule may contain indirect links.

Note how the orange homomorphism in the bottom left does not match to the newly created edges, nor does the green homomorphism from D.

The purple homomorphism from L does match the indirect link onto the created edges. But the red homomorphism does not match onto the indirect links in L, as they do not exist in K.

\bentley{Fix this up a lot.}

\subsubsection*{DSLTrans Layer Execution}

Now that the production of a particular DSLTrans rule has been explained, the next definition defines how a DSLTrans transformation layer executes. Recall that a layer is composed of rules, and acts on the current input-output model.

\begin{definition} {DSLTrans Layer Execution}
\label{def:dsltrans_layer_execution}

The execution of a DSLTrans layer can be defined as a function as follows:

\textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM

Let $l = \{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$, $l \in \textsc{Layers}$.

Let $\mathit{IOM}_{Input}$, $\mathit{IOM}_{Output} \in \mathit{IOMs}$.

$\mathit{IOM}_{Output}$ will be the result of applying the production \textit{p} as defined below to $\mathit{IOM}_{Input}.$

We refer to the double-pushout literature for the creation of \textit{p}. Specifically, we refer to the definition of parallel direct derivations. This p will be created out of the productions created for each of the rules $r_i$ in the layer.

$p = \big\langle (p_1, {in}^1), \dots, (p_k, {in}^k) \big\rangle : (L \xleftarrow{l} L \xrightarrow{r} R)$

This collection of the rules is possible because DSLTrans rules do not delete anything. Essentially, a union is taken of the L, K, R components of each production.
\end{definition}

\begin{figure}[h!] \centering
\includegraphics[width=0.44\textwidth]{figures/formal/parallel_productions}
	\caption{Parallel production \textit{p}}
	\label{fig:parallel_productions}
\end{figure}
